%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
%                                                                 %
%                            CHAPTER                              %
%                                                                 %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 

\chapter{Conclusion and future research}
\label{ch:Conclusions}

This chapter will contain some finals words on the problem definition and the proposed implementation, as well as some recommendations for future research.

\section{Conclusion}

//TODO

\paragraph{Problem definition} 

The first step in analysing DNA, is to sequence it. This will determine which bases and in which order they are occuring in the DNA. However, the technology of DNA sequencing machines limit us to reads of 75-300 bases long. Therefore, the DNA is multiplied and cut into short pieces before sequencing. If we want to make assumptions about the whole genome, we will need to input a lot of these reads. 

Typically, the read is compared with the whole genome in a local alignment, for example with the Smith-Waterman algorithm. As an output, we would get the position in the human genome and an alignment with its score (how well the sequence fits in that spot). This practice is commonly referred to as \emph{Mapping to reference genome}.

Since the reads from DNA sequencing machines are 75 to 300 bases long, and the whole human genome is approximately 3 billion bases, this comparison is computationally a very intensive task. If we analyze the S-W algorithm (as we have done in \ref{expl:SWanalyse}), we can see that the value of each cell in the matrix is only dependent on the left-upmost 3 cells. Therefore, It leads us to believe that this algorithm can be accelerated on other hardware solutions such as an FPGA (as discussed in \ref{ch:Platforms}) since S-W is heavily parallelizable.

In most clinical applications where mapping to a human reference genome is used, the number of reads to be compared with the genome is in the millions, which further increases the demand to speedup the process of mapping the reads in the genome, to decrease the time-to-result.

The idea of this thesis was to implement the Smith-Waterman alignment algorithm on an MPSoC, which contains both programmable hardware and an ARM processor. As a target board, the ZCU 104 evaluation kit was chosen.

\paragraph{Proposed solution and implementation}

As a starting point, a software implementation was implemented on the target board.

\paragraph{Speedup}



\section{Recommendations for future work}

Naturally, we have several recommendations for the continuation of this research. They will be classified in improvements of the current implementation and re-evaluation of the used algorithm.

\subsection{Recommendations for improving the current implementation}

Before, during, and after implementing, some optimizations which could improve the proposed implementation came up. However, we did not have time to implement them, so hereunder they will be mentioned:

\begin{itemize}
	\item Treat the gap opening and extension separately. In the current implementation, the gap penalty is treated linearly. However, the S-W algorithm could be refined to include an affine gap penalty. More info on the gap penalty can be found at \ref{expl:SWanalyse}.
	\item In a sequence, a base may be suddenly marked with an 'N' character, Which represents a base that was unidentified in the primary processing. In this thesis, it was decided to cut the sequence short at the moment an 'N' character is registered. However, a possible improvement might be to keep these unidentified bases in the alignment as a "generic" base and in this way improve the accuracy of the alignment. More info on the current implementation can be found at \ref{codeStructure}.
	\item Make some types more memory efficient, such as the BASE and the SEQ\_INDEX type. Firstly, the BASE type only has 4 possible values (the 4 nucleotide bases). However, it is defined as 1 byte in the current implementation, since it is the smallest type that C supports. However, storing a base in 1 byte is memory inefficient. Improvements could be made by defining an own type, by stacking 4 bases in 1 byte. Secondly, the SEQ\_INDEX type is stored as 2 bytes in the current implementation, even though it only contains values up to 300, which makes it memory inefficient. This could also be improved.
	\item If this implementation would be used in practice, there would be a need to find an easier way to on and offload data to board. Currently,  this is accomplished by changing out the SD card, which could easily be improved using FTP since an Ethernet stack is present due to the operating system. The speed the data would be transferred at, is not important since we can assume the time it takes to map the sequence takes far longer. 
\end{itemize}

\subsection{Recommendations for re-evaluating the alignment method}

In the field of alignment algorithms, a few alternatives exist to S-W since it takes a lot of computing power. In most cases, these algorithms will use a transformation or a lookup table to determine possible candidate locations, where a small regional S-W will be performed on a part of the genome. For example, The BFAST algorithm uses a hash table where candidate locations are stored for every sequence \cite{fpgaImpl}. As a second example, in the Bowtie application, the candidate locations are determined using the Burrows-Wheeler Transform.

As future work, one of these more sophisticated algorithms can be selected. Some candidate locations can then be determined in a way defined by the algorithm. These candidate locations can then be mapped using the implementation of this thesis.